Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-07-09 MS/ms.Rmd
adding a box (not done yet)

We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms then including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are known to differ. By synthesizing diverse study outcomes, it allows us to not just report on overall effects (if the question is relevant to ask), but to understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction) and approaches that can explain the diversity of effect outcomes observed. In doing so, it provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included within a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogenity estimates then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogenity estimates, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant, and indeed may be misleading. Therefore, understanding overall effects in each of these groups seperately (and whether they differ) will be more useful. It is important to consider these questions when conducting meta-analyses.

@@

#### *Comparing Changes in Mean Physiological Rates, $Q_{10}$*
A very common occurrence in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using normal effect size measures (see above) would result in the effect sizes varying in the types of temperatures applied within and across across studies contributing to 'comparability' issues. As such, it is often of interest to standardize effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

@@

Note that this formula does not include $\sigma^2_{m}$ as sampling error variance is assumed to be known in meta-analysis as explained above. When including continuous moderators, such as dosage, temperature, salinity, and even pH, $R^2_{marginal}$ provides a formal means to assess just how much variation in effects is the result of 'nuisance' heterogeneity at the effect size level. Other biological moderators of interest could also be included as fixed effects and $R^2_{marginal}$ calculated for each independently or all together. Importantly, moderators could be those that explain variation at the effect size level (i.e., dosage, temperature) or they could explain variation at the study or species level (e.g., reproductive mode, endothermy etc). $R^2_{marginal}$, or how we like to think of it, $R^2_{'nuisance'}$, maybe a useful measure to help readers understand just how much effect variation can result from the specific treatment application. If the heterogeneity is high as a result of, say, dosages differences, then it is clear that the choice of dosage within and across studies will have a critical impact on the effect of interest. Comparative physiologists intersted in implementing thse calculations can do so using our packages `orchaRd` [@Nakagawa2021c] or `metaAidR` (https://github.com/daniel1noble/metaAidR).

Heterogenity described using $I^2$ measures can be useful for understanding the relative contributions of different factors to effect size variation, however, prediction intervals might be more appropriate in many cases [@Nakagawa2021c; @Borenstein2019]. Unlike $I^2$, prediction intervals (PI's) are probably more meaningful in the context of meta-analysis because they explicitly incorperate measures of dispersion around a mean effect size providing information about the likely effect size one can expect if we were to randomly sample a new population. For example, assume that we were interested in the overall impact of salinity stress in freshwater fish. To tackle this question, we collected experimental studies measuring freshwater fish swim performance under high salinity treatments (~15-20ppm NaCl) compared to freshwater (~0ppm NaCl) controls and used SMD to quantify the magnitude of effect salinity had on swim performance. We conduct a meta-analysis and estimate an overall SMD of 0.50 with a prediction interval of 0.05 to 0.85. The PI indicates that effects can vary widely from as low as an SMD of 0.05 to as high as an SMD 0.95, depending on the population. In otherwords, if we were to repeat a study we would expect a new effect to fall within the range of 0.05 and 0.85, 95\% of the time [@Borenstein2019; @Nakagawa2021c]. This is a very intutive interprettaion of how variable effects are, yet PIs are often not reported in meta-analyses [@Nakagawa2021c].

## **Conclusions and Future Directions**
Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental approaches / designs and environments. While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, obvious drivers of effect size heterogeneity, such as temperature and dosage differences across studies, may not really be of interest. Indeed, comparative physiologists are likely to want an effect size or analytical approach that allows them to remove these 'nuisances'. Here, we provide comparative physiologists with a set of tools to deal with 'nuisance heterogeneity' by creatively 'reinventing' standard effect sizes or using meta-regression models. Of course, nuisance heterogeneity might in of itself be the main interest, in which case standard meta-regression approaches will work sufficiently. However, in many cases, comparative physiologists might simply want to focus on bigger, more interesting biological drivers of effect size direction and magnitude. However, estimating complex meta-regression models with a multitude of fixed effects might result in more challenging model interpretation. As such, a combined approach, where some of the known heterogeneity is dealt with by deriving a variant effect size and associated sampling variance, and other biologically interesting moderators are fit in a meta-regression model, might be the desired. Indeed, this may even be a necessity to simplify the model. We hope the set of tools we describe here provide clarification around sourfes of heterogenity and provides ways to help comparative physiologists to communicate effect measures more richly to help guide our understanding and decisions around pressing global challenges.

>>>>

We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms then including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are known to differ. By synthesizing diverse study outcomes, it allows us to not just report on overall effects (if the question is relevant to ask), but to understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction) and approaches that can explain the diversity of effect outcomes observed. In doing so, it provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included within a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogeneity estimates then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogeneity estimates, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant, and indeed may be misleading. Therefore, understanding overall effects in each of these groups separately (and whether they differ) will be more useful. It is important to consider these questions when conducting meta-analyses.

@@

#### *Comparing Changes in Mean Physiological Rates,* $Q_{10}$

A very common occurrence in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using normal effect size measures (see above) would result in the effect sizes varying in the types of temperatures applied within and across across studies contributing to 'comparability' issues. As such, it is often of interest to standardize effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

@@

Note that this formula does not include $\sigma^2_{m}$ as sampling error variance is assumed to be known in meta-analysis as explained above. When including continuous moderators, such as dosage, temperature, salinity, and even pH, $R^2_{marginal}$ provides a formal means to assess just how much variation in effects is the result of 'nuisance' heterogeneity at the effect size level. Other biological moderators of interest could also be included as fixed effects and $R^2_{marginal}$ calculated for each independently or all together. Importantly, moderators could be those that explain variation at the effect size level (i.e., dosage, temperature) or they could explain variation at the study or species level (e.g., reproductive mode, endothermy etc). $R^2_{marginal}$, or how we like to think of it, $R^2_{'nuisance'}$, maybe a useful measure to help readers understand just how much effect variation can result from the specific treatment application. If the heterogeneity is high as a result of, say, dosages differences, then it is clear that the choice of dosage within and across studies will have a critical impact on the effect of interest. Comparative physiologists intersted in implementing thse calculations can do so using our packages `orchaRd` [@Nakagawa2021c] or `metaAidR` (<https://github.com/daniel1noble/metaAidR>).

Heterogenity described using $I^2$ measures can be useful for understanding the relative contributions of different factors to effect size variation, however, prediction intervals might be more appropriate in many cases [@Nakagawa2021c; @Borenstein2019]. Unlike $I^2$, prediction intervals (PI's) are probably more meaningful in the context of meta-analysis because they explicitly incorperate measures of dispersion around a mean effect size providing information about the likely effect size one can expect if we were to randomly sample a new population. For example, assume that we were interested in the overall impact of salinity stress in freshwater fish. To tackle this question, we collected experimental studies measuring freshwater fish swim performance under high salinity treatments (\~15-20ppm NaCl) compared to freshwater (\~0ppm NaCl) controls and used SMD to quantify the magnitude of effect salinity had on swim performance. We conduct a meta-analysis and estimate an overall SMD of 0.50 with a prediction interval of 0.05 to 0.85. The PI indicates that effects can vary widely from as low as an SMD of 0.05 to as high as an SMD 0.95, depending on the population. In otherwords, if we were to repeat a study we would expect a new effect to fall within the range of 0.05 and 0.85, 95% of the time [@Borenstein2019; @Nakagawa2021c]. This is a very intutive interprettaion of how variable effects are, yet PIs are often not reported in meta-analyses [@Nakagawa2021c].

## **Conclusions and Future Directions**

Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental approaches / designs and environments. While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, obvious drivers of effect size heterogeneity, such as temperature and dosage differences across studies, may not really be of interest. Indeed, comparative physiologists are likely to want an effect size or analytical approach that allows them to remove these 'nuisances'. Here, we provide comparative physiologists with a set of tools to deal with 'nuisance heterogeneity' by creatively 'reinventing' standard effect sizes or using meta-regression models. Of course, nuisance heterogeneity might in of itself be the main interest, in which case standard meta-regression approaches will work sufficiently. However, in many cases, comparative physiologists might simply want to focus on bigger, more interesting biological drivers of effect size direction and magnitude. However, estimating complex meta-regression models with a multitude of fixed effects might result in more challenging model interpretation. As such, a combined approach, where some of the known heterogeneity is dealt with by deriving a variant effect size and associated sampling variance, and other biologically interesting moderators are fit in a meta-regression model, might be the desired. Indeed, this may even be a necessity to simplify the model. We hope the set of tools we describe here provide clarification around sourfes of heterogenity and provides ways to help comparative physiologists to communicate effect measures more richly to help guide our understanding and decisions around pressing global challenges.
